---
title: "PCA Outcomes"
author: "Xin (David) Zhao"
date: "`r Sys.Date()`"
output: md_document
params:
    group_1: "WT-NC"
    group_2: "WT-HFD"
    host: "Mice"
    order: 1
    pca_comp1: "30%"
    pca_comp2: "12%"
    f_value: "0.85366"  
    r_squared: "0.078652"
    p_value: "0.457"
--- 




```{r setup, include = FALSE} 

library(dplyr)

group1 <- params$group_1
group2 <- params$group_2 

# File paths to scores plots 
scores_plot_path <- paste0("../data/Download_",group1, "_", group2, "/pca_score2d_0_dpi72.png")   

# File paths to loading tables  
loading_path <- paste0("../data/Download_",group1, "_", group2, "/pca_loadings.csv")  

# File paths to metabolite names files 
meta_name_path <- paste0("../data/Download_",group1, "_", group2, "/TMIC-02RX-metabolite-names.csv") 

# File paths to exported csv 
exported_file_path <- paste0("../data/Download_",group1, "_", group2, "/pca_loadings2.csv")  


# Iteration order 
iteration_order <- params$order 

```


```{r results='asis', echo=FALSE}
if (iteration_order == 1) {
    cat("\n## PCA\n")
} else {
    cat("\n")
}
```


### `r params$group_1` vs `r params$group_2`  

**Scores Plot of PCA**

This figure shows a Principal Component Analysis (PCA) scores plot illustrating the variations in metabolite profiles of mice under `r params$group_1` and `r params$group_2` groups. Each point represents a sample, with clusters indicating separation or overlap between the two groups based on their metabolic profiles.

The first and second principal components explain approximately `r params$pca_comp1` and `r params$pca_comp2` of the total variation in metabolites for mice under `r params$group_1` and `r params$group_2` groups.

Each point represents a sample, with colors indicating different groups. The PERMANOVA test shows no significant difference in centroids between `r params$group_1` and `r params$group_2` groups (PERMANOVA F-value: `r params$f_value`; R-squared: `r params$r_squared`; p-value (based on 999 permutations): `r params$p_value`).

```{r, echo=FALSE, fig.cap='', out.width='85%', fig.align='center'}

knitr::include_graphics(scores_plot_path)

```

**Figure 5. PCA Scores Plot - Metabolite Profile Variations in `r params$host` under `r params$group_1` and `r params$group_2` groups**

**Loadings of PCA**

This table lists the PCA loadings for the top 20 metabolites contributing to the variation in metabolite profiles of mice under `r params$group_1` and `r params$group_2` groups. Higher loadings indicate greater influence of a metabolite on the principal components, highlighting the most significant contributors to the observed metabolic differences between the two groups.

The original, full loading table can be found in the appendix, `r loading_path` 

```{r echo=FALSE}

pca_loadings_wtwt <- read.csv(loading_path)  

names(pca_loadings_wtwt)[1] <- "metabolite" 
meta_full_name <- read.csv(meta_name_path)  
pca_loadings_wtwt <- pca_loadings_wtwt %>% left_join(meta_full_name[ ,c(1,2)], by="metabolite")  
pca_loadings_wtwt <- pca_loadings_wtwt %>% select(metabolite, class, everything())  
# Subset PC1-5 
pca_loadings_wtwt2 <- pca_loadings_wtwt |> dplyr::select(metabolite: PC5) 

# Order metabolites from large to small by PC1 absolute values 
pca_loadings_wtwt2 <- pca_loadings_wtwt2[order(abs(pca_loadings_wtwt2$PC1), decreasing = TRUE), ]

# Rename X to Metabolite 
colnames(pca_loadings_wtwt2)[1] <- "Metabolites" 

# Save as csv file 
write.csv(pca_loadings_wtwt2, exported_file_path, row.names = FALSE) 

# Top metabolites only 
pca_loadings_wtwt3 <- pca_loadings_wtwt2 |> head(20)

knitr::kable(pca_loadings_wtwt3, digits = 3, format = 'simple', align = 'l', row.names = FALSE, caption = '')

```

**Table 1. PCA Loadings for Metabolite Profile Variations in Mice under `r params$group_1` and `r params$group_2` groups**

